---
title: "Coral reef causal inference review"
author: "Jess Prevost"
date: "`r Sys.Date()`"
bibliography: causal_inf_citations.bib
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Review

### Overarching question - are their scientific conclusions justified by their data and analysis?

### How well defined is the causal question?

The main causal question is outlined in the abstract and revisited throughout the text. The question is defined as how did specific key variables causally affect the recovery trajectories of coral reefs in the Seychelles following bleaching events in 1998 [@graham_2015]. Since causal inference is the prediction of intervention [@pearl_2009], the causal question is predicting whether changes in any of the key variables would change the recovery trajectory of a reef. The question is fairly specific indicating that the results apply only to these reefs in the Seychelles and this specific bleaching event. They do not generalise to different areas although the authors do use the defined causal model, to predict the recovery trajectory of the same reefs which had since recovered following a bleaching event in 2016. 

Some detail in the variables were omitted from the paper but these were mostly available in the research by @graham_2015. The alternative recovery trajectories of coral reefs were defined as recovering or regime shifting as per @graham_2015. Recovering reefs had coral cover greater than macroalgal coverage and time series data showed high or increasing coral cover while regime shifting reefs showed macroalgae dominance greater than coral cover and macroalgal cover remained high or increased over time [@graham_2015]. Four metrics were used to define recovery trajectories with the full methodology described in @graham_2015. The year of data collection was stated as 1994 for branching coral, macroalgal cover, herbivorous fish biomass, and structural complexity to account for these variables prior to bleaching and the other variables collection dates were not stated but can be found in [@graham_2015].
There was however a discrepancy relating the the data collection for the nutrient measurements. @arif_2022 stated that these data were from 2004 and suggested @graham_2015 as the source of the data but according to @graham_2015 nutrient measurements were collected in 2014. This may have been a typographical error since the same values for nutrients are used in the causal model dataset and prediction dataset and it is stated that the predictive model uses data from 2014. The 2014 data from @graham_2015 is unfortunately unavailable thus it cannot be cross-checked. If this is not a typographical error, the source of the nutrient data should have been cited or data collection described. Either way, the nutrient measurements were taken after the bleaching event occurred. Nutrient content is expected to change over seasons but differences in sites are anticipated to remain stable [@graham_2015]. Nevertheless, there is the possibility for the post-bleaching macroalgal content to affect nutrient levels which would be an issue for inference from measurements taken pre-bleaching. Depth, MPA, and wave exposure data were expected to remain stable over time thus even though depth was measured and wave exposure was calculated post-bleaching (1998 - 2011), this should not be an issue for inference as macroalgae content post-bleaching should not affect depth, MPA, or wave exposure. 

The specification of the causal model as a DAG was good, using domain knowledge and extensive research. Each causal relationship in the DAG was detailed in the supplementary text explaining why each of the causal relationships were included. No potentially unobserved variables included. There is a risk of using only variables which are available/ have been collected.  
Potential for branching coral to cause MPA selection (higher preservation of diversity).

DAG-data consistency was tested for using the d-separation method [@pearl_2009]. All d-separation conditions implied by DAG were consistent with the data.

There was discrepancy between the DAG downloaded from dagitty.net in the supplied code and the DAG stated in the paper [@arif_2022]. The downloaded DAG contained a directed edge from depth to initial macroalgae cover and excluded the directed edge from wave exposure to branching coral unlike the DAG in the paper. Nonetheless, the causal relationships described in the supplementary information and the d-separation conditions examined were associated with the DAG in the paper. After checks it was found that the data were in fact consistent with the d-separation conditions implied by both DAGs. However, this does raise the question of whether the final causal model stated in the paper was the first to be proposed or whether alternative DAGs, such as the one in the supplied code, were tested. The authors recognise that multiple DAGs can pass DAG-data consistency, and that the final DAG should be justified with theory but do not outline their consideration of alternative DAGs. Ideally, a consideration of alternative DAGs and explanation of why the final causal model was chosen should have been included, especially since the data were consistent with both of these DAGs. 


### The causal adjustment strategy
The authors used the backdoor criterion to define an adjustment set for each variable in order to estimate the total causal effect of the given variable on the recovery trajectory of coral reefs. This is an established method for calculating sets of variables to adjust for to give an unbiased estimate of the total causal effect of one variable on another [@pearl_2009]. 
All the models specified in the paper satisfy the backdoor criterion for the DAG used in the paper. The model specifications for some of the variables would change if the DAG in the supplied code were used instead (such as branching coral). For the majority of variables including MPA, depth, wave exposure, nutrients, branching, structural complexity, and herbivore biomass, the minimal adjustment set was used for the model specifications, meaning the set containing the fewest variables which satisfies the backdoor criterion. However, for the initial macroalgae model, depth was included as a covariate even though the minimal adjustment set does not include this variable. The adjustment set still satisfies the backdoor criterion with depth included, it is just not the minimal adjustment set. The inclusion of depth might be as a result of the use of the DAG in the supplied code prior to the final DAG included in the paper. The supplementary text hints at this with the inclusion of paths 5 and 6 which do not exist in the DAG in the paper but do exist in the DAG in the supplied code. The directed edge from depth to initial macroalgae opens additional backdoor paths which should be adjusted for by including depth when using the DAG in the supplied code. Reasons for the decision to use the minimal adjustment sets for the other variables and not to use the minimal adjustment set for the initial macroalgae model should have been explicitly stated. While all adjustment sets which satisfy the backdoor criterion give unbiased estimates of the total causal effect but they can differ in precision of the estimates depending on the additional variables included [@pearl_2009]. 

There were additional errors in the supplementary information for this section (S1.3). The initial macroalgae model had an incomplete path (path 4) listed, although this did not affect the adjustment set. The model specified in the paper for herbivore biomass which adjusts for structural complexity and MPA is the minimial set satisfying the backdoor criterion. However, the supplementary text description has an error stating that structural complexity and branching coral should be adjusted for. This is incorrect as this is insufficient as an adjustment set since branching coral is a collider and opens another backdoor path (Herbivore biomass <- MPA -> Branching coral <- Depth -> Regime Shift) thus MPA or Depth would still additionally need to be included in the set to block this newly opened path. 


### Bayesian analysis

priors seem good, based on literature
only used single chain
rhat not stated nor convergence acknowedged
no seed set - not reproducible

table S2 - no indication of what [,] are - looks like 90% interval based on fig 3a

### potential limitations of the study, reported or otherwise
data collection biases

the measurements in the causal model were taken in 1994 for some of the variables which is 2 years prior to the bleaching event - could have changed in this time period

other causes - substrate type? also check @graham_2015

In predictive model, was updated macroalgal cover considered (could some reefs that recovered be weaker than others due to previous bleachings)

For predictive model, different rates of recovery after other bleaching events could be a new cause


rightly interpretted only the effect size of the predicotr variable of interest

### any other aspects that you think are important for causal inference

Question the change in causal model - it seems to be retrospective as the code and some details in the supplementary info were not changed - why was this the case? were the estimates better? was it during review?

MPA - binary variable? not stated 

losing information by using a binary response

Grammatical/typesetting issues
Few errors in beta numbers in statistical model specifications in paper and supplementary (branching coral model, inital macroalgae model)
Supplementary text 
Missing a heading for branching coral model specification in S1.3

Having the code and data definitely makes the analysis more reproducible 

# Reproduction
In order to reproduce the analysis by @arif_2022, I obtained the data and R code supplied from Figshare [@arif_2021] and modified the code in R to attempt to reproduce the results [@R]. (As alluded to above, the metadata for the supplied data was poor with no description for any of the four datasets nor the variables within them other than what could be drawn from the paper, the supplementary data, and the code.) I read in the four datasets and checked for missing values.

I did not reproduce the creation of the DAG as this requires extensive knowledge about coral reef systems, and the reasons for the directed relationships within the DAG are outlined well in the supplementary text. After reading in the DAG from dagitty.web using the *dagitty* package [@dagitty], I noted that the DAG in the code differed from that in the paper as previously mentioned. Thus, I modified the DAG layout code to reflect the DAG in the paper. I confirmed that there were 32 DAG-implied independencies, cross-checked them against those listed in the supplementary data, and confirmed that the observed data were consistent with the independencies (p = 1) after performing Holm-Bonferroni correction to p-values to account for multiple testing. To assess the selection of covariates in the models for each variable, I drew the DAG and used the backdoor criterion to determine which variables to adjust for. I also made use of the adjustmentSets function to determine the minimal adjustment set for each of the variables [@dagitty]. As noted above, the minimal adjustment set was used for all variables except the initial macroalgae model. For the remainder of the analysis, I made use of the models stated in the paper since these all still satisfied the backdoor criterion. 

As in the analysis by @arif_2022, I used the *rethinking* package [@rethinking] to fit Bayesian logistic regression models for each of the variables of interest with the response being 0 for recovering sites and 1 for sites showing a shift in regime. The data were standardised by subtracting the mean of each variable and dividing by two standard deviations prior to model fitting. I made use of the newer ulam tool rather than map2stan for model fitting [@rethinking]. Weakly informative priors were used including a Cauchy(0,10) prior for the intercept terms and student_t(4,0,2.5) priors for slope coefficients as recommended by previous authors. 


```{r packages, include = FALSE}
library(pacman)
p_load(readxl, dagitty, tidyverse, rstanarm, rstan, rethinking, ggeffects)

# Install cmdstanr and rethinking
# install.packages(c("coda","mvtnorm","devtools","loo","dagitty","shape"))
# devtools::install_github("rmcelreath/rethinking")
```

```{r read_data}

# Standardised data - this was used in analysis
SeychellesS <- read_xlsx("data/SeychellesS.xlsx") %>%
  mutate(across(c(Region, habitat), as.factor),
         across(c(MPA, Depth, Branching, Struccomplex, Nutrients, Herb, Macro...16), as.numeric))
str(SeychellesS)
summary(SeychellesS)
any(is.na(SeychellesS))

# Unstandardised data
Seychelles <- read_xlsx("data/Seychelles.xlsx")

# 2014 observational dataset for prediction
Seychelles2014 <- read_xlsx("data/Seychelles2014.xlsx") 
Predictions <- read_xlsx("data/Predictions.xlsx")

```

```{r DAG}
# Test DAG-data consistency

# Download DAG from dagitty.net
DAG <- downloadGraph("dagitty.net/mbt1AvL") 

# Visualise DAG
plot(graphLayout(DAG))

# View conditional independencies
print(impliedConditionalIndependencies(DAG))
length(impliedConditionalIndependencies(DAG))

# Evaluate the d-separation implications of the DAG
r <- localTests(x = DAG, data = SeychellesS %>% 
                  select(MPA, Branching, Struccomplex, Herb, Shifted, Depth, Nutrients, Macro...16, Wave) %>%
                  rename(Nutrient = Nutrients, Macro = Macro...16))
r

# Perform Holm-Bonferroni correction to p-values
r$p.value <- p.adjust(r$p.value)
r
# p-values all 1 indicate DAG-data consistency 

# View minimal adjustment sets when using the backdoor criterion (i.e. effect = "total")
adjustmentSets(DAG, exposure = "MPA", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Nutrient", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Herb", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Herb", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Branching", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Depth", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Struccomplex", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Wave", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Macro", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Macro", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG, exposure = "Macro", outcome = "Shifted", effect = "total") 

# Obtain code for DAG
cat(as.character(DAG))

# Edit code to reflect DAG in paper
DAG2 <- dagitty(
'dag {
bb="0,0,1,1"
Branching -> Shifted
Branching -> Struccomplex
Depth -> Branching
Depth -> Nutrient
Depth -> Shifted
Herb -> Macro
Herb -> Shifted
MPA -> Branching
MPA -> Herb
Macro -> Shifted
Nutrient -> Macro
Nutrient -> Shifted
Struccomplex -> Herb
Struccomplex -> Shifted
Wave -> Branching
Wave -> Macro
Wave -> Shifted
}'
)

# Visualise DAG
plot(graphLayout(DAG2))

# View conditional independencies
print(impliedConditionalIndependencies(DAG2))
length(impliedConditionalIndependencies(DAG2))

# Evaluate the d-separation implications of the DAG
r2 <- localTests(x = DAG2, data = SeychellesS %>% 
                  select(MPA, Branching, Struccomplex, Herb, Shifted, Depth, Nutrients, Macro...16, Wave) %>%
                  rename(Nutrient = Nutrients, Macro = Macro...16))
r2

# Perform Holm-Bonferroni correction to p-values
r2$p.value <- p.adjust(r2$p.value)
r2
# p-values all 1 indicate DAG-data consistency 

# View minimal adjustment sets when using the backdoor criterion (i.e. effect = "total")
adjustmentSets(DAG2, exposure = "MPA", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Nutrient", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Herb", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Branching", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Depth", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Struccomplex", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Wave", outcome = "Shifted", effect = "total") 
adjustmentSets(DAG2, exposure = "Macro", outcome = "Shifted", effect = "total") 

# View all possible adjustment sets for initial macroalgae when using the backdoor criterion 
adjustmentSets(DAG2, exposure = "Macro", outcome = "Shifted", effect = "total", type = "all") 

```

```{r logistic_regression, include = F}
# First fit logistic regression models just to check 
summary(SeychellesS)
glm1 <- glm(Shifted ~ MPA, data = SeychellesS)
summary(glm1)
```

```{r bayesian_models}
# Use ulam with cmdstan rather than map2stan

# Set seed for reproducibility
set.seed(123)

# Shifted ~ bernoulli(p)

#--------------------------------
## MPA model: 
# logit(p) = beta0 + beta1 * MPA

MPA_mod <- ulam(
  flist = # flist contains the formulas that define the likelihood and priors
    alist(Shifted ~ dbinom(1, p), # shifted has bernoulli distribution
          logit(p) <- beta0 + beta1 * MPA, # logistic model
          beta0 ~ dcauchy(0, 10), # priors on beta0
          beta1 ~ dstudent(4, 0, 2.5) # priors on beta1
    ), 
  data = SeychellesS %>% select(Shifted, MPA), 
  cmdstan = TRUE)

precis(MPA_mod) # summary statistics for posteriors
plot(MPA_mod) # visual representations of summary statistics
rethinking::traceplot(MPA_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(MPA_mod)
post_MPA <- post$beta1
PI90_MPA <- PI(post_MPA, prob = 0.90)
PI50_MPA <- PI(post_MPA, prob = 0.50) 
med_MPA <- median(post_MPA) 
str(SeychellesS)

#--------------------------------
## Nutrient model: 
# logit(p) = beta0 + beta1 * Nutrients + beta2 * Depth

nutrient_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Nutrients + beta2 * Depth,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5),
          beta2 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Nutrients, Depth), 
  cmdstan = TRUE)

precis(nutrient_mod) # summary statistics for posteriors
plot(nutrient_mod) # visual representations of summary statistics
rethinking::traceplot(nutrient_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(nutrient_mod)
post_nutrient <- post$beta1
PI90_nutrient <- PI(post_nutrient, prob = 0.90)
PI50_nutrient <- PI(post_nutrient, prob = 0.50) 
med_nutrient <- median(post_nutrient) 

#--------------------------------
## Herbivore biomass model:
# logit(p) = beta0 + beta1 * Herb + beta2 * MPA + beta3 * Struccomplex

herb_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Herb + beta2 * MPA + beta3 * Struccomplex ,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5),
          beta2 ~ dstudent(4, 0, 2.5),
          beta3 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Herb, MPA, Struccomplex), 
  cmdstan = TRUE)

precis(herb_mod) # summary statistics for posteriors
plot(herb_mod) # visual representations of summary statistics
rethinking::traceplot(herb_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(herb_mod)
post_herb <- post$beta1
PI90_herb <- PI(post_herb, prob = 0.90)
PI50_herb <- PI(post_herb, prob = 0.50) 
med_herb <- median(post_herb) 


#--------------------------------
## Branching coral model:
# logit(p) = beta0 + beta1 * Branching + beta2 * MPA + beta3 * Depth + beta4 * Wave

branching_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Branching + beta2 * MPA + beta3 * Depth + beta4 * Wave,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5),
          beta2 ~ dstudent(4, 0, 2.5),
          beta3 ~ dstudent(4, 0, 2.5),
          beta4 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Branching, MPA, Depth, Wave), 
  cmdstan = TRUE)

precis(branching_mod) # summary statistics for posteriors
plot(branching_mod) # visual representations of summary statistics
rethinking::traceplot(branching_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(branching_mod)
post_branching <- post$beta1
PI90_branching <- PI(post_branching, prob = 0.90)
PI50_branching <- PI(post_branching, prob = 0.50) 
med_branching <- median(post_branching) 

#--------------------------------
## Depth model:
# logit(p) = beta0 + beta1 * Depth

depth_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Depth,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Depth), 
  cmdstan = TRUE)

precis(depth_mod) # summary statistics for posteriors
plot(depth_mod) # visual representations of summary statistics
rethinking::traceplot(depth_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(depth_mod)
post_depth <- post$beta1
PI90_depth <- PI(post_depth, prob = 0.90)
PI50_depth <- PI(post_depth, prob = 0.50) 
med_depth <- median(post_depth) 

#--------------------------------
## Structural complexity model:
# logit(p) = beta0 + beta1 * Struccomplex + beta2 * Branching

complexity_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Struccomplex + beta2 * Branching,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5),
          beta2 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Struccomplex, Branching), 
  cmdstan = TRUE)

precis(complexity_mod) # summary statistics for posteriors
plot(complexity_mod) # visual representations of summary statistics
rethinking::traceplot(complexity_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(complexity_mod)
post_complexity <- post$beta1
PI90_complexity <- PI(post_complexity, prob = 0.90)
PI50_complexity <- PI(post_complexity, prob = 0.50) 
med_complexity <- median(post_complexity) 

#--------------------------------
## Wave model:
# logit(p) = beta0 + beta1 * Wave

wave_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Wave,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Wave), 
  cmdstan = TRUE)

precis(wave_mod) # summary statistics for posteriors
plot(wave_mod) # visual representations of summary statistics
rethinking::traceplot(wave_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(wave_mod)
post_wave <- post$beta1
PI90_wave <- PI(post_wave, prob = 0.90)
PI50_wave <- PI(post_wave, prob = 0.50) 
med_wave <- median(post_wave) 

#--------------------------------
## Initial macroalgae biomass model:
# logit(p) = beta0 + beta1 * Macro...16 + beta2 * Herb + beta3 * Depth + beta4 * Nutrients + beta5 * Wave

macro_mod <- ulam(
  flist = 
    alist(Shifted ~ dbinom(1, p),
          logit(p) <- beta0 + beta1 * Macro + beta2 * Herb + beta3 * Depth + beta4 * Nutrients + beta5 * Wave,
          beta0 ~ dcauchy(0, 10),
          beta1 ~ dstudent(4, 0, 2.5),
          beta2 ~ dstudent(4, 0, 2.5),
          beta3 ~ dstudent(4, 0, 2.5),
          beta4 ~ dstudent(4, 0, 2.5),
          beta5 ~ dstudent(4, 0, 2.5)
    ), 
  data = SeychellesS %>% select(Shifted, Macro...16, Herb, Depth, Nutrients, Wave) %>%
  rename(Macro = Macro...16), 
  cmdstan = TRUE)

precis(macro_mod) # summary statistics for posteriors
plot(macro_mod) # visual representations of summary statistics
rethinking::traceplot(macro_mod) # traceplots

# show median and 90% and 50% percentile interval (PI) for depth  
post <- extract.samples(macro_mod)
post_macro <- post$beta1
PI90_macro <- PI(post_macro, prob = 0.90)
PI50_macro <- PI(post_macro, prob = 0.50) 
med_macro <- median(post_macro) 

```

```{r fig_3A}
# Use median and 50% and 90% PI values from above models to create Manuscript Figure 3a. 

label <- c("MPA", "Carbon:Nitrogen", "Herbivore Biomass", "Branching Coral", "Depth", "Structural Complexity", "Wave Exposure", "Initial Macroalgae")
median <- c(med_MPA, med_nutrient, med_herb, med_branching, med_depth, med_complexity, med_wave, med_macro)

lower90 <- c(PI90_MPA[1], PI90_nutrient[1], PI90_herb[1], PI90_branching[1], PI90_depth[1], PI90_complexity[1], PI90_wave[1], PI90_macro[1])
upper90 <- c(PI90_MPA[2], PI90_nutrient[2], PI90_herb[2], PI90_branching[2], PI90_depth[2], PI90_complexity[2], PI90_wave[2], PI90_macro[2])

lower50 <- c(PI50_MPA[1], PI50_nutrient[1], PI50_herb[1], PI50_branching[1], PI50_depth[1], PI50_complexity[1], PI50_wave[1], PI50_macro[1])
upper50 <- c(PI50_MPA[2], PI50_nutrient[2], PI50_herb[2], PI50_branching[2], PI50_depth[2], PI50_complexity[2], PI50_wave[2], PI50_macro[2])

df <- data.frame(label, median, lower90, upper90, lower50, upper50)

# reverses the factor level ordering for labels after coord_flip()
df$label <- factor(df$label, levels=rev(df$label))

ggplot(data=df, aes(x=label, y=median, ymin=lower90, ymax=upper90)) +
  geom_errorbar(aes(ymin=lower50, ymax=upper50), width =0, size = 1.3)+ # I just added a new error bar on top of your original error bar with the dummy data that I made (lower50, upper50). width = 0 gets ride of the whiskers, size controls the thickness. you can also do color = "put your favoriate color" if you wan to use color 
  geom_pointrange() + 
  geom_hline(yintercept=0, lty=2) +  # add a dotted line at x=0 after flip
  coord_flip() +  # flip coordinates (puts labels on y axis)
  xlab(" ") + ylab("Standardized Effect Size") +
  theme_bw()  # use a white background
```
```{r}
# Plot marginal plots of continuous predictor variables as shown in Figure 3b and 4b. Use ‘Seychelles’ dataset, which has unstandardized data.

# Depth marginal plot 
fit <- stan_glm(Shifted ~ Depth, data = Seychelles, family = binomial("logit"),chains = 1)
depth_MP <- ggpredict(fit)
plot(depth_MP, show_data = TRUE, jitter = 0)

# Nutrient marginal plot 
fit <- stan_glm(Shifted ~ Nutrients + Depth, data = Seychelles, family = binomial("logit"), chains = 1)
nutrient_MP <- ggpredict(fit)
plot(nutrient_MP, show_data = TRUE, jitter = 0.0)

# Branching coral marginal plot
# Fix this - MPA is not in dataset (take from SeychellesS and unstandardise)
fit <- stan_glm(Shifted ~ Branching + Depth + MPA + Wave, data = Seychelles, family = binomial("logit"), chains = 1)
branching_MP <- ggpredict(fit)
plot(branching_MP, show_data = TRUE, jitter = 0.0)

# Structural complexity marginal plot 
fit <- stan_glm(Shifted ~ Struccomplex + Branching, data = Seychelles, family = binomial("logit"),chains = 1)
struccomplex_MP <- ggpredict(fit) 
plot(struccomplex_MP, show_data = TRUE, jitter = 0.0)

# Wave exposure marginal plot 
fit <- stan_glm(Shifted ~ Wave, data = Seychelles, family = binomial("logit"), chains = 1)
wave_MP <- ggpredict(fit)
plot(wave_MP, show_data = TRUE, jitter = 0.0)

# Herbivore biomass marginal plot 
fit <- stan_glm(Shifted ~ Herb + Struccomplex + Branching, data = Seychelles, family = binomial("logit"), chains = 1)
herb_MP <- ggpredict(fit)
plot(herb_MP, show_data = TRUE, jitter = 0.0)

# Initial macroalgae cover marginal plot 
fit <- stan_glm(Shifted ~ Macro + Herb + Depth + Nutrients + Wave, data = Seychelles, family = binomial("logit"), chains = 1)
macro_MP <- ggpredict(fit, terms = "Macroalgae[0:20 by=0.1]")
plot(macro_MP, show_data = TRUE, jitter = 0.0)

```



